### 2018-03-26 Maths Notes/Further Pure/FP2/Maclaurin and Taylor Series/Maclaurin and Taylor Series.tex
Notes up to date

\begin{document}
\begin{center}
\underline{\huge Maclaurin and Taylor Series}
\end{center}
\section{Maclaurin's expansion}
$$f(x)=f(0)+f'(0)x+f''(0)\frac{x^2}{2!}+f'''(0)\frac{x^3}{3!}+...+f^r(0)\frac{x^r}{r!}...$$
\\
For the continuous function, f, given by $f:x\Rightarrow f(x)$(where $x$ is real), then providing $f(0),f'(0),f''(0)$ etc all have finite values. This is an infinite series.
\subsection{Example}
\textit{Given that $f(x)=e^x$ can be written as an infinite series in the form:}
$$f(x)=e^x=a_0+a_1x+a_2x^2+a_3x^3+...+a_rx^4+...$$
\textit{And that it is valid to differentiate an infinite series term by term, show that:}
$$e^2=1+x+\frac{x^2}{2!}+\frac{x^3}{3!}+...+\frac{x^r}{r!}+...$$
\\
\\
\textcolor{red}{Find up to the third differential of $f(x)$ and the value of zero for each}\\
\begin{tabular}{l l}
$f(x)=e^x$&$f(0)=1$\\
$f'(x)=e^x$&$f''(0)=1$\\
$f''(x)=e^x$&$f''(0)=1$\\
$f'''(x)=e^x$&$f'''(0)=1$
\end{tabular}
\\
\\
$$f(x)=1+1\times x+1\times\frac{x^2}{2!}+1\times\frac{x^3}{3!}$$
$$f(x)=1+x+\frac{x^2}{2!}+\frac{x^3}{3!}+...+\frac{x^r}{r!}+...$$
\subsection{Standard results}
Standard results are given on the data sheet, these can then be used for adapted forms of the results also. Remember to consider the limits where appropriate.
\newpage
\section{Taylor expansion}
The conditions of the Maclaurin expansion mean that some functions, such as $\ln x$ cannot be expanded as a series in ascending powers of x.\\
\\
The construction of the Maclaurin expansion focuses on $x=0$ and values of x very close to zero. The Taylor expansion focuses on $x=a$.\\
\\
Considering the functions f and g, where $f(x+a)\equiv g(x)$ then:\\
$f^r(a)=g^r(0)$\\
\\
Turning the Maclaurin expansion for g from:
$$g(x)=g(0)+g'(0)x+\frac{g''(0)}{2!}x^2...$$
Into
$$f(x+a)=f(a)+f'(a)x+\frac{f''(a)}{2!}x^2+...+\frac{f^r(a)}{r!}x^r$$
Replacing x by x-a gives
$$f(x)=f(a)+f'(a)(x-a)+\frac{f''(a)}{2!}(x-a)^2+...+\frac{f^r(a)}{r!}(x-a)^r$$
These are the two forms of the Taylor expansion, when a=0, they both become the Maclaurin expansion.
\end{document}

>>>>

\begin{document}
\begin{center}
\underline{\huge Maclaurin and Taylor Series}
\end{center}
\section{Maclaurin's expansion}
$$f(x)=f(0)+f'(0)x+f''(0)\frac{x^2}{2!}+f'''(0)\frac{x^3}{3!}+...+f^r(0)\frac{x^r}{r!}...$$
\\
For the continuous function, f, given by $f:x\Rightarrow f(x)$(where $x$ is real), then providing $f(0),f'(0),f''(0)$ etc all have finite values. This is an infinite series.
\subsection{Example}
\textit{Given that $f(x)=e^x$ can be written as an infinite series in the form:}
$$f(x)=e^x=a_0+a_1x+a_2x^2+a_3x^3+...+a_rx^4+...$$
\textit{And that it is valid to differentiate an infinite series term by term, show that:}
$$e^2=1+x+\frac{x^2}{2!}+\frac{x^3}{3!}+...+\frac{x^r}{r!}+...$$
\\
\\
\textcolor{red}{Find up to the third differential of $f(x)$ and the value of zero for each}\\
\begin{tabular}{l l}
$f(x)=e^x$&$f(0)=1$\\
$f'(x)=e^x$&$f''(0)=1$\\
$f''(x)=e^x$&$f''(0)=1$\\
$f'''(x)=e^x$&$f'''(0)=1$
\end{tabular}
\\
\\
$$f(x)=1+1\times x+1\times\frac{x^2}{2!}+1\times\frac{x^3}{3!}$$
$$f(x)=1+x+\frac{x^2}{2!}+\frac{x^3}{3!}+...+\frac{x^r}{r!}+...$$
\subsection{Standard results}
Standard results are given on the data sheet, these can then be used for adapted forms of the results also. Remember to consider the limits where appropriate.
\newpage
\section{Taylor expansion}
The conditions of the Maclaurin expansion mean that some functions, such as $\ln x$ cannot be expanded as a series in ascending powers of x.\\
\\
The construction of the Maclaurin expansion focuses on $x=0$ and values of x very close to zero. The Taylor expansion focuses on $x=a$.\\
\\
Considering the functions f and g, where $f(x+a)\equiv g(x)$ then:\\
$f^r(a)=g^r(0)$\\
\\
Turning the Maclaurin expansion for g from:
$$g(x)=g(0)+g'(0)x+\frac{g''(0)}{2!}x^2...$$
Into
$$f(x+a)=f(a)+f'(a)x+\frac{f''(a)}{2!}x^2+...+\frac{f^r(a)}{r!}x^r$$
Replacing x by x-a gives
$$f(x)=f(a)+f'(a)(x-a)+\frac{f''(a)}{2!}(x-a)^2+...+\frac{f^r(a)}{r!}(x-a)^r$$
These are the two forms of the Taylor expansion, when a=0, they both become the Maclaurin expansion.
\subsection{Example}
\textit{Find the Taylor expansion of $\cos2x$, in ascending powers of $(x-\frac{\pi}{4})$ up to $(x-\frac{\pi}{4})^5$}
\textcolor{red}{Find the differentials of $f(x)$ up to the fifth derivative, and the associated values when $x=\frac{\pi}{4}$}\\
\begin{tabular}{l l}
$f(x)=\cos2x$&$f(\frac{\pi}{4})=0$\\
$f'(x)=-2\sin2x$&$f(\frac{\pi}{4})=-2$\\
$f''(x)=-4\cos2x$&$f(\frac{\pi}{4})=0$\\
$f'''(x)=8\sin2x$&$f(\frac{\pi}{4})=8$\\
$f''''(x)=16\cos2x$&$f(\frac{\pi}{4})=0$\\
$f'''''(x)=-32\sin2x$&$f(\frac{\pi}{4})=-32$
\end{tabular}\\
\textcolor{red}{Substitute in the associated values into the formula}
$$\cos2x=0-2\Big(x-\frac{\pi}{4}\Big)+0+\frac{8}{3!}\Big(x-\frac{\pi}{4}\Big)^3+0-\frac{32}{5!}\Big(x-\frac{\pi}{4}\Big)^5$$
$$\cos2x=-2\Big(x-\frac{\pi}{4}\Big)+\frac{4}{5}\Big(x-\frac{\pi}{4}\Big)^3-\frac{4}{15}\Big(x-\frac{\pi}{4}\Big)^5$$

\newpage
\section{Finding the solution, in the form of a series to a differential equation using the Taylor series method}
Suppose you have a first order differential equation of the form $\frac{dy}{dx}=f(x,y)$ and you know the initial condition that at $x=x_0$,$y=y_0$, then you can calculate \Large{$(\frac{dy}{dx})_{x_0}$} \normalsize by substituting $x_0$ and $y_0$ into the original differential equation.\\
\\
By successive differentiation of the original differential equation, the values of {\Large $(\frac{d^2y}{dx^2})_{x_0}$} and {\Large $(\frac{d^3y}{dx^3})_{x_0}$} and so on can be found by substituting previous results into the derived equations.\\
\\
The series solution to the differential equation is found using the Taylor series in the form:
$$y=y_0+(x-x_0)\bigg(\frac{dy}{dx}\bigg)_{x_0}+\frac{(x-x_0)^2}{2!}\Bigg(\frac{d^2y}{dx^2}\Bigg)_{x_0}+\frac{(x-x_0)^3}{3!}\Bigg(\frac{d^3y}{dx^3}\Bigg)_{x_0}+...$$
In the common situation where $x_0=0$ then this reduces to the Maclaurin series
$$y=y_0+x\bigg(\frac{dy}{dx}\bigg)_{x_0}+\frac{x^2}{2!}\Bigg(\frac{d^2y}{dx^2}\Bigg)_{x_0}+\frac{x^3}{3!}\Bigg(\frac{d^3y}{dx^3}\Bigg)_{x_0}+...$$
\subsection{Example}
\textit{Using the Taylor method to find a series solution, in ascending powers of x up to and including the term in $x^3$, of:}
$$\frac{d^2y}{dx^2}=y-\sin x$$
\textit{Given that when $x=0$,$y=1$ and $\frac{dy}{dx}=2$}
\\
\textcolor{red}{Use the formula to find $\frac{d^2y}{dx^2}$}
$$\frac{d^2y}{dx^2}=1-\sin(0)=1$$
$$\mathbf{\frac{d^2y}{dx^2}=1}$$
\textcolor{red}{Differentiate the formula to obtain a formula for $\frac{d^3y}{dx^3}$}
$$\frac{d^3y}{dx^3}=\frac{dy}{dx}-\cos x=2-\cos(0)=1$$
$$\mathbf{\frac{d^3y}{dx^3}=1}$$
\textcolor{red}{Substitute the known values into the Maclaurin formula}
$$y=1+x\times2+\frac{x^2}{2!}\times1+\frac{x^3}{3!}\times1$$
$$y=1+2x+\frac{x^2}{2}+\frac{x^3}{6}$$
\end{document}

### 2018-03-26 Physics Notes/Turning Points/Special relativity/Special Relativity.tex
Notes up to date

\begin{document}
\begin{center}
\underline{\huge Special Relativity}
\end{center}
\section{Frames of reference}
Einstein's theory of special relativity is the result of analysing the consequences of the absence of any universal frame of reference. Special relativity concentrates on frames of reference which are moving at a constant velocity relative to each other. These frames are known as \textbf{inertial frames of reference}.\\
\\
Special relativity is based on two postulates:
\begin{itemize}
\item The laws of physics, expressed in equations, have the same form in all inertial frames
\item The speed of light in free space is invariant (ie the same for all observers regardless of their state of motion and the speed of the light source)
\end{itemize}
\section{The Michelson-Morley experiment}
The experiment was carried out to find out if the direction of the Earth's motion through space differs from the speed of light in a perpendicular direction. This was investigated using the Michelson interferometer, shown below:\\
\includegraphics[width=10cm]{michelson.png}\\
The semi-silvered glass block splits the beam of monochromatic light into two beams at A which is on the silvered side of the block. The block, $P_2$, is necessary to ensure that both beams pass through the same thickness of glass and air. If the speed of light depended on the angle compared to the earth's rotation there would be interference at O due to the waves no longer being in phase. However this was not detected, meaning there is no difference in speed.
\section{Time dilation and proper time}
A consequence of the invariance of the speed of light in free space is that time runs slower when in motion. \\
\includegraphics[width=8cm]{time_dilation.png}\\
This shows that to an observer time is slower to an observer.\\
\\
$$t=\frac{t_0}{\sqrt{1-(\frac{v}{c})^2}}$$
t- Time measured by external observer\\
$t_0$ - Time measured in the moving frame (called the proper time)\\
v - Relative velocity of moving frame\\
c- Speed of light in a vacuum
\subsection{Example experiment}
When cosmic rays reach the ionosphere above the Earth's surface, unstable sub-atomic particles called muons are created, travelling at 0.966c. These particles have a half life of $1.5\mu s$.\\
\\
Measurements by Bruno Rossi showed that about 80\% of the muons created by cosmic radiation on a mountain top reached an observatory 2km below. In one half-life of $1.5\mu s$, muons moving at 0.996c would decay, resulting in fewer detected at the bottom than the top.
\section{Length contraction}
A moving observer sees an object to appear smaller than it is (as measured in the stationary frame).\\
\\
Time contraction is \textbf{only} along the direction of travel.\\
It s easier to think that a moving observer measures the distance they cover to be less than a stationary observer.
$$L=L_0\sqrt{1-(\frac{v}{c})^2}$$
L - Length measured by the moving observer\\
$L_0$ - Length measured by the stationary observer (called the proper length)\\
v- Relative velocity of the moving frame\\
c- Speed of light in a vacuum
\end{document}

>>>>

\begin{document}
\begin{center}
\underline{\huge Special Relativity}
\end{center}
\section{Frames of reference}
Einstein's theory of special relativity is the result of analysing the consequences of the absence of any universal frame of reference. Special relativity concentrates on frames of reference which are moving at a constant velocity relative to each other. These frames are known as \textbf{inertial frames of reference}.\\
\\
Special relativity is based on two postulates:
\begin{itemize}
\item The laws of physics, expressed in equations, have the same form in all inertial frames
\item The speed of light in free space is invariant (ie the same for all observers regardless of their state of motion and the speed of the light source)
\end{itemize}
\section{The Michelson-Morley experiment}
The experiment was carried out to find out if the direction of the Earth's motion through space differs from the speed of light in a perpendicular direction. This was investigated using the Michelson interferometer, shown below:\\
\includegraphics[width=10cm]{michelson.png}\\
The semi-silvered glass block splits the beam of monochromatic light into two beams at A which is on the silvered side of the block. The block, $P_2$, is necessary to ensure that both beams pass through the same thickness of glass and air. If the speed of light depended on the angle compared to the earth's rotation there would be interference at O due to the waves no longer being in phase. However this was not detected, meaning there is no difference in speed.
\section{Time dilation and proper time}
A consequence of the invariance of the speed of light in free space is that time runs slower when in motion. \\
\includegraphics[width=8cm]{time_dilation.png}\\
This shows that to an observer time is slower to an observer.\\
\\
$$t=\frac{t_0}{\sqrt{1-(\frac{v}{c})^2}}$$
t- Time measured by external observer\\
$t_0$ - Time measured in the moving frame (called the proper time)\\
v - Relative velocity of moving frame\\
c- Speed of light in a vacuum
\subsection{Example experiment}
When cosmic rays reach the ionosphere above the Earth's surface, unstable sub-atomic particles called muons are created, travelling at 0.966c. These particles have a half life of $1.5\mu s$.\\
\\
Measurements by Bruno Rossi showed that about 80\% of the muons created by cosmic radiation on a mountain top reached an observatory 2km below. In one half-life of $1.5\mu s$, muons moving at 0.996c would decay, resulting in fewer detected at the bottom than the top.
\section{Length contraction}
A moving observer sees an object to appear smaller than it is (as measured in the stationary frame).\\
\\
Time contraction is \textbf{only} along the direction of travel.\\
It s easier to think that a moving observer measures the distance they cover to be less than a stationary observer.
$$L=L_0\sqrt{1-(\frac{v}{c})^2}$$
L - Length measured by the moving observer\\
$L_0$ - Length measured by the stationary observer (called the proper length)\\
v- Relative velocity of the moving frame\\
c- Speed of light in a vacuum
\section{Relativistic mass, momentum and energy}
\subsection{Relativistic mass}
Rest mass (proper mass,$m_0$):Mass as measured by an observer who is stationary relative to the mass\\
Relative mass(m): The apparent mass of an object that is moving relative to the observer (speed v)\\
$$m=\frac{m_0}{\sqrt{1-(\frac{v}{c})^2}}$$
Relativistic mass increases with speed
\newpage
\subsection{Relativistic momentum}
$$p=mv-\frac{m_0v}{\sqrt{1-(\frac{v}{c})^2}}$$
\subsection{Relativistic energy}
$$E=mc^2=\frac{m_0c^2}{\sqrt{1-(\frac{v}{c})^2}}$$
\subsubsection{Kinetic energy}
\begin{center}
Total energy=Rest energy+Relativistic KE
\end{center}
$$E=E_0+E_K$$
$$E_K=E-E_0$$
$$E_K=\frac{m_0c^2}{\sqrt{1-(\frac{v}{c})^2}}-m_0c^2$$
$$E_K=m_0c^2\Bigg(\bigg(1-\bigg(\frac{v}{c}\bigg)^2\bigg)^{-\frac{1}{2}}-1\Bigg)$$
\textcolor{red}{Apply binomial expansion to the first term}
$$E_K=m_0c^2\Bigg(1+\frac{1}{2}\Big(\frac{v}{c}\Big)^2+\frac{-\frac{1}{2}(-\frac{1}{2}-1)}{2}\times\Big(\frac{v}{c}\Big)^4-1\Bigg)$$
\textcolor{red}{Approximate away all terms larger than $\Big(\frac{v}{c}\Big)^2$ and simplify}
$$E_K\approx m_0c^2\times\frac{1}{2}\times\frac{v^2}{c^2}=\frac{1}{2}m_0v^2$$
Together mass and energy are conserved, but not necessaries individually
\section{Summary of equations}
\begin{tabular}{l l}
Gamma factor&$\gamma=\sqrt{1-(\frac{v}{c})^2}$\\
Length contraction&$L_0\gamma$\\
Time dilation&\Large{$\frac{t_0}{\gamma}$}\\
Relativistic mass&\Large{$\frac{m_0}{\gamma}$}\\
Relativistic momentum&\Large{$\frac{m_0v}{\gamma}$}\\
Relativistic energy&\Large{$\frac{m_0c^2}{\gamma}$}
\end{tabular}
\end{document}
